---
title: "Project in Statistics"
author: "Marie Ramm"
format: html
editor: visual
---

Project in Statistics

Marie Ramm, jrl918

June 2026

University of Copenhagen

Advisor: Predrag Pilipovic

# Library

```{r}
# ---- LOAD PACKAGES ---- 
library(dplyr)
library(skimr)
library(tibble)
library(grid)
library(gridExtra)
library(ggplot2)
library(zoo)
library(tidyr)
library(expm) 
library(httr)
library(readxl)
library(dplyr)
library(patchwork)


```

# Data Preprocessing

```{r}
excel_url = "https://www.iceandclimate.nbi.ku.dk/data/GICC05modelext_GRIP_and_GISP2_and_resampled_data_series_Seierstad_et_al._2014_version_10Dec2014-2.xlsx"
res = httr::GET(excel_url)
excel_file = tempfile(fileext = ".xlsx")
# Write downloaded content to file
writeBin(content(res, "raw"), excel_file)

Ca2data = readxl::read_excel(excel_file, sheet = 3,
                              range = "A51:I12278") %>%
  slice(-1) %>%
  select(1, ncol(.)) %>%
  mutate_all(as.numeric) %>%
  rename(age = 1, Ca2 = 2)
unlink(excel_file)
readr::write_csv(Ca2data, excel_file)
```

```{r}
# Active df in time-frame 7900 - 104000, NA filled w/ linear interpolation, 4804 rows
Ca2_final = Ca2data %>%
  filter(age >= 30000, age < 80000) %>%
  mutate(Ca2 = na.approx(Ca2)) %>%
  group_by(age) %>%
  summarise(Ca2 = mean(Ca2), .groups = "drop") %>%
  mutate(Ca2 = -log(Ca2)) %>%
  mutate(X = Ca2 - mean(Ca2)) %>%
  mutate(age = age/1000) %>%
  arrange(desc(age))  %>% 
  mutate(V = (lead(X) - X)/0.02) %>% 
  filter(is.finite(V))

Ca2_data = cbind(Ca2_final$X, Ca2_final$V)
```

```{r}
p1 = ggplot(Ca2_final, aes(x = age, y = X)) +
  geom_line(linewidth = 0.7) +
  labs(
    title = expression(X[t]),
    x = "Age (kiloyears before present)",
    y = expression(X[t])
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank() 
  ) +
  scale_x_reverse()

p2 = ggplot(Ca2_final, aes(x = age, y = V)) +
  geom_line(linewidth = 0.7) +
  labs(
    title = expression(V[t]),
    x = "Age (kiloyears before present)",
    y = expression(V[t])
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  scale_x_reverse()

grid.arrange(
  p1,
  p2,
  ncol = 1, 
  top = expression("Transformed Greenland Ice-Core Process " * X[t] * " and Estimated Velocity " * V[t])
)
```

```{r}
p1Density = ggplot(Ca2_final, aes(x = X)) +
  geom_density() +
  labs(
    title = expression("Ice Land Core Greenland: X[t] Density"),
    x = "Density",
    y = expression(X[t])
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank() 
  ) +
  scale_x_reverse()

p2Density = ggplot(Ca2_final, aes(x = V)) +
  geom_density() +
  labs(
    title = expression("V[t] Density"),
    x = "Density",
    y = expression(V[t])
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  scale_x_reverse()

grid.arrange(
  p1Density,
  p2Density,
  ncol = 1
)
```

```{r}
base_theme = theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  )

# ---- Top left: X_t over time ----
p_xt = ggplot(Ca2_final, aes(x = age, y = X)) +
  geom_line(linewidth = 0.4) +
  # optional equilibrium lines:
  # geom_hline(yintercept = c(eq1, eq2), colour = "darkred", linewidth = 0.8) +
  labs(
    x = NULL,
    y = expression(X[t])
  ) +
  scale_x_reverse() +
  base_theme +
  theme(
    axis.title.x = element_blank(),
    plot.margin = margin(5.5, 5.5, 2, 5.5)
  )

# ---- Bottom left: V_t over time ----
p_vt = ggplot(Ca2_final, aes(x = age, y = V)) +
  geom_line(linewidth = 0.35) +
  labs(
    x = "time before present [ka]",
    y = expression(V[t])
    # If using rescaled forward differences instead:
    # y = expression(sqrt(3)/2 * (X[t+h] - X[t])/h)
  ) +
  scale_x_reverse() +
  base_theme +
  theme(
    plot.margin = margin(2, 5.5, 5.5, 5.5)
  )

# ---- Top right: density of X_t ----
p_xdens = ggplot(Ca2_final, aes(x = X)) +
  geom_density(linewidth = 0.8, colour = "black", fill = "darkred", alpha = 0.5) +
  # optional equilibrium lines:
  # geom_vline(xintercept = c(eq1, eq2), colour = "darkred", linewidth = 0.8) +
  coord_flip() +
  labs(
    x = NULL,
    y = NULL
  ) +
  base_theme +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks = element_blank(),
    plot.margin = margin(5.5, 5.5, 2, 2)
  )

# ---- Bottom right: density of V_t ----
p_vdens = ggplot(Ca2_final, aes(x = V)) +
  geom_density(linewidth = 0.8, colour = "black", fill = "darkred", alpha = 0.5) +
  coord_flip() +
  labs(
    x = "empirical density",
    y = NULL
  ) +
  base_theme +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.margin = margin(2, 5.5, 5.5, 2)
  )

((p_xt | p_xdens) / (p_vt | p_vdens)) +
    plot_layout(widths = c(2, 1), heights = c(1, 1)) +
    plot_annotation(
        title = "Greenland Ice-Core Data: Process and Empirical Densities",
        theme = theme(
            plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
        )
    )
```

# M0 - constant diffusion, full

## Likelihood

### Rough

```{r}
neg_loglik = function(theta, data, h) {
  
  # ---- sort data ---- 
  data_old = data[-nrow(data), ]
  data_new = data[-1, ]
  n = nrow(data_old)
  
  # ---- Moments ---- #  generally move up 
  X_old = data_old[,1]
  Xexp  = mean(X_old)
  Xexp2 = mean(X_old^2)
  varX  = var(X_old)

  #---- theta ----
  eta   = theta[1]
  a     = theta[2]
  b     = theta[3]
  c     = theta[4]
  d     = theta[5]
  gamma = theta[6]
  
  drift_const = 3 * a * Xexp2 + 2 * b * Xexp + c

  A = matrix(c(0, 1,
                drift_const, -eta),
              nrow = 2, byrow = TRUE)
  
  # ---- B ----
  inside_sqrt = varX + (Xexp + b / (3 * a))^2

  bxPlus = -b / (3 * a) + sqrt(inside_sqrt)
  bxMinus = -b / (3 * a) - sqrt(inside_sqrt)

  B = cbind(ifelse(X_old > 0, bxPlus, bxMinus),0)

    # ---- f ---- 
  f_step = function(df, h) {
        
        X = df[,1]
        V = df[,2]
        
        NonLin =
            a * X^3 +
            b * X^2 +
            c * X +
            d -
            drift_const * (X -  B[, 1])
        
        out = cbind(
            X,
            V + h * NonLin
        )
        
        return(out)
    }
  
  # ---- Mu Step ---- 
  mu_step = function(Y, df) {
        
        expAh = expm(A * h)

        mu = as.matrix(Y - B) %*% t(expAh) + B
        
        return(mu)
  }
  
  # ---- Residual ---- 
  Y_old_half = f_step(data_old, h = 1/2*h)
  Y_new_inv  = f_step(data_new, h = -1/2*h)
    
  Y_mu = mu_step(Y_old_half, data_old)
    
  Z = Y_new_inv - Y_mu
    
  # ---- Constant diffusion covariance ----
  Sigma = matrix(
      c(0, 0,
        0, gamma),
      nrow = 2,
      byrow = TRUE
    )
    
   M = rbind(
       cbind(A, Sigma),
       cbind(matrix(0, 2, 2), -t(A))
    )
   
  R = expm(M * h)
    
  RBias = expm(M*(3/2)*h)
  
  I = R[1:2, 3:4]
  F1 = R[1:2, 1:2]
    
  Omega = I %*% t(F1)
  
    
  # Rough
  
  OmegaBias = (RBias[1:2, 3:4] %*% t(RBias[1:2, 1:2]))[2,2]
  
  log_det = log(OmegaBias)

  first = Z[,2]^2 * 1/(Omega[2, 2])

  val = sum(first) + n*(2/3*log_det)
  
 return(val)
 }

```

### Full

```{r}
neg_loglik_full = function(theta, data, h) {
  
  # ---- sort data ---- 
  data_old = data[-nrow(data), ]
  data_new = data[-1, ]
  n = nrow(data_old)
  
  # ---- Moments ---- 
  X_old = data_old[,1]
  Xexp  = mean(X_old)
  Xexp2 = mean(X_old^2)
  varX  = var(X_old)

  #---- theta ----
  eta   = theta[1]
  a     = theta[2]
  b     = theta[3]
  c     = theta[4]
  d     = theta[5]
  gamma = theta[6]
  
  drift_const = 3 * a * Xexp2 + 2 * b * Xexp + c

  A = matrix(c(0, 1,
                drift_const, -eta),
              nrow = 2, byrow = TRUE)
  
  # ---- B ----
  inside_sqrt = varX + (Xexp + b / (3 * a))^2

  bxPlus = -b / (3 * a) + sqrt(inside_sqrt)
  bxMinus = -b / (3 * a) - sqrt(inside_sqrt)

  B = cbind(ifelse(X_old > 0, bxPlus, bxMinus),0)

    # ---- f ---- 
  f_step = function(df, h) {
        
        X = df[,1]
        V = df[,2]
        
        NonLin =
            a * X^3 +
            b * X^2 +
            c * X +
            d -
            drift_const * (X -  B[, 1])
        
        out = cbind(
            X,
            V + h * NonLin
        )
        
        return(out)
    }
  
  # ---- Mu Step ---- 
  mu_step = function(Y, df) {
        
        expAh = expm(A * h)

        mu = as.matrix(Y - B) %*% t(expAh) + B
        
        return(mu)
  }
  
  # ---- Residual ---- 
  Y_old_half = f_step(data_old, h = 1/2*h)
  Y_new_inv  = f_step(data_new, h = -1/2*h)
    
  Y_mu = mu_step(Y_old_half, data_old)
    
  Z = Y_new_inv - Y_mu
    
  # ---- Constant diffusion covariance ----
  Sigma = matrix(
      c(0, 0,
        0, gamma),
      nrow = 2,
      byrow = TRUE
    )
    
   M = rbind(
       cbind(A, Sigma),
       cbind(matrix(0, 2, 2), -t(A))
    )
   
  R = expm(M * h)
    
  RBias = expm(M*(3/4)*h)
  
  I = R[1:2, 3:4]
  F1 = R[1:2, 1:2] 
    
  Omega = I %*% t(F1)
  
  #FULL 
   
  OmegaBias =  RBias[1:2, 3:4] %*% t(RBias[1:2, 1:2])

  omega_inv = solve(Omega)

  log_det = log(det(OmegaBias))

  first = Z[,1]^2 * omega_inv[1, 1] +
    2 * Z[,1] * Z[,2] * omega_inv[1, 2] +
    Z[,2]^2 * omega_inv[2, 2]


  val = sum(first) + n*(4/3*log_det)
    
 return(val)
 }

```

### Full no bias

```{r}
neg_loglik_no_bias = function(theta, data, h) {
  
  # ---- sort data ---- 
  data_old = data[-nrow(data), ]
  data_new = data[-1, ]
  n = nrow(data_old)
  
  # ---- Moments ---- 
  X_old = data_old[,1]
  Xexp  = mean(X_old)
  Xexp2 = mean(X_old^2)
  varX  = var(X_old)

  #---- theta ----
  eta   = theta[1]
  a     = theta[2]
  b     = theta[3]
  c     = theta[4]
  d     = theta[5]
  gamma = theta[6]
  
  drift_const = 3 * a * Xexp2 + 2 * b * Xexp + c

  A = matrix(c(0, 1,
                drift_const, -eta),
              nrow = 2, byrow = TRUE)
  
  # ---- B ----
  inside_sqrt = varX + (Xexp + b / (3 * a))^2

  bxPlus = -b / (3 * a) + sqrt(inside_sqrt)
  bxMinus = -b / (3 * a) - sqrt(inside_sqrt)

  B = cbind(ifelse(X_old > 0, bxPlus, bxMinus),0)

    # ---- f ---- 
  f_step = function(df, h) {
        
        X = df[,1]
        V = df[,2]
        
        NonLin =
            a * X^3 +
            b * X^2 +
            c * X +
            d -
            drift_const * (X -  B[, 1])
        
        out = cbind(
            X,
            V + h * NonLin
        )
        
        return(out)
    }
  
  # ---- Mu Step ---- 
  mu_step = function(Y, df) {
        
        expAh = expm(A * h)

        mu = as.matrix(Y - B) %*% t(expAh) + B
        
        return(mu)
  }
  
  # ---- Residual ---- 
  Y_old_half = f_step(data_old, h = 1/2*h)
  Y_new_inv  = f_step(data_new, h = -1/2*h)
    
  Y_mu = mu_step(Y_old_half, data_old)
    
  Z = Y_new_inv - Y_mu
    
  # ---- Constant diffusion covariance ----
  Sigma = matrix(
      c(0, 0,
        0, gamma),
      nrow = 2,
      byrow = TRUE
    )
    
   M = rbind(
       cbind(A, Sigma),
       cbind(matrix(0, 2, 2), -t(A))
    )
   
  R = expm(M * h)
  
  I = R[1:2, 3:4]
  F1 = R[1:2, 1:2] 
    
  Omega = I %*% t(F1)
  
  #FULL 
  omega_inv = solve(Omega)

  log_det = log(det(Omega))

  first = Z[,1]^2 * omega_inv[1, 1] +
    2 * Z[,1] * Z[,2] * omega_inv[1, 2] +
    Z[,2]^2 * omega_inv[2, 2]


  val = sum(first) + n*(log_det)
    
 return(val)
 }
```

## Optimize

```{r}
theta_start0 = c(
  eta = 60,
  a = -130,   
  b = 0,
  c = 115,
  d = 0,
  gamma = 8000 
)

fitBFGS = optim(
  par = theta_start0,
  fn = neg_loglik,
  data = Ca2_data,
  h = 0.02,
  method = "L-BFGS-B",
  lower = c(
    eta = 1e-8,
    a = -Inf,
    b = -Inf,
    c = -Inf,
    d = -Inf,
    gamma = 200
  ),
  upper = c(
    eta = Inf,
    a = -1e-8,
    b = Inf,
    c = Inf,
    d = Inf,
    gamma = Inf
  )
)

# ---- BFGS ---- 
fitBFGS$value
fitBFGS$convergence
fitBFGS$par
```

```{r}
fitBFGS_full = optim(
  par = theta_start0,
  fn = neg_loglik_full,
  data = Ca2_data,
  h = 0.02,
  method = "L-BFGS-B",
  lower = c(
    eta = 1e-8,
    a = -Inf,
    b = -Inf,
    c = -Inf,
    d = -Inf,
    gamma = 200
  ),
  upper = c(
    eta = Inf,
    a = -1e-8,
    b = Inf,
    c = Inf,
    d = Inf,
    gamma = Inf
  )
)

# ---- BFGS ---- 
fitBFGS_full$value
fitBFGS_full$convergence
fitBFGS_full$par
```

```{r}
fitBFGS_no_bias = optim(
  par = theta_start0,
  fn = neg_loglik_no_bias,
  data = Ca2_data,
  h = 0.02,
  method = "L-BFGS-B",
  lower = c(
    eta = 1e-8,
    a = -Inf,
    b = -Inf,
    c = -Inf,
    d = -Inf,
    gamma = 200
  ),
  upper = c(
    eta = Inf,
    a = -1e-8,
    b = Inf,
    c = Inf,
    d = Inf,
    gamma = Inf
  )
)

# ---- BFGS ---- 
fitBFGS_no_bias$value
fitBFGS_no_bias$convergence
fitBFGS_no_bias$par
```

# M1 - constant diffusion, b=d=0

## Likelihood

### Rough

```{r}
neg_loglik1 = function(theta, data, h) {
  
  # ---- sort data ---- 
  data_old = data[-nrow(data), ]
  data_new = data[-1, ]
  n = nrow(data_old)
  
  # ---- Moments ---- 
  X_old = data_old[,1]
  Xexp  = mean(X_old)
  Xexp2 = mean(X_old^2)
  varX  = var(X_old)

  #---- theta ----
  eta   = theta[1]
  a     = theta[2]
  c     = theta[3]
  gamma = theta[4]
  b = 0 
  d = 0
  
  drift_const = 3 * a * Xexp2 + 2 * b * Xexp + c

  A = matrix(c(0, 1,
                drift_const, -eta),
              nrow = 2, byrow = TRUE)
  
  # ---- B ----
  inside_sqrt = varX + (Xexp + b / (3 * a))^2

  bxPlus = -b / (3 * a) + sqrt(inside_sqrt)
  bxMinus = -b / (3 * a) - sqrt(inside_sqrt)

  B = cbind(ifelse(X_old > 0, bxPlus, bxMinus),0)

  # ---- f ---- 
  f_step = function(df, h) {
        
        X = df[,1]
        V = df[,2]
        
        NonLin =
            a * X^3 +
            b * X^2 +
            c * X +
            d -
            drift_const * (X -  B[, 1])
        
        out = cbind(
            X,
            V + h * NonLin
        )
        
        return(out)
    }
  
  # ---- Mu Step ---- 
  mu_step = function(Y, df) {
        
        expAh = expm(A * h)

        mu = as.matrix(Y - B) %*% t(expAh) + B
        
        return(mu)
  }
  
  # ---- Residual ---- 
  Y_old_half = f_step(data_old, h = 1/2*h)
  Y_new_inv  = f_step(data_new, h = -1/2*h)
    
  Y_mu = mu_step(Y_old_half, data_old)
    
  Z = Y_new_inv - Y_mu
    
  # ---- Constant diffusion covariance ----
  Sigma = matrix(
      c(0, 0,
        0, gamma),
      nrow = 2,
      byrow = TRUE
    )
    
   M = rbind(
       cbind(A, Sigma),
       cbind(matrix(0, 2, 2), -t(A))
    )
   
  R = expm(M * h)
    
  RBias = expm(M*(3/2)*h)
  
  I = R[1:2, 3:4]
  F1 = R[1:2, 1:2]
    
  Omega = I %*% t(F1)
  
  # FULL 
   
  # OmegaBias =  RBias[1:2, 3:4] %*% t(RBias[1:2, 1:2])
  # 
  # omega_inv = solve(Omega)
  #   
  # log_det = as.numeric(determinant(OmegaBias, logarithm = TRUE)$modulus)
  #   
  # first = Z[,1]^2 * omega_inv[1, 1] +
  #   2 * Z[,1] * Z[,2] * omega_inv[1, 2] +
  #   Z[,2]^2 * omega_inv[2, 2] 
  # 
  #   
  # val = sum(first) + n*(4/3*log_det)
    
  # Rough
  
  OmegaBias = (RBias[1:2, 3:4] %*% t(RBias[1:2, 1:2]))[2,2]

  #omega_inv = solve(Omega)
    
  # log_det = as.numeric(determinant(OmegaBias, logarithm = TRUE)$modulus)
  
  log_det = log(OmegaBias)
    
  # first = Z[,1]^2 * omega_inv[1, 1] +
  #   2 * Z[,1] * Z[,2] * omega_inv[1, 2] +
  #   Z[,2]^2 * omega_inv[2, 2] 
  
  first = Z[,2]^2 * 1/(Omega[2, 2])

  val = sum(first) + n*(2/3*log_det)
  
 return(val)
 }

```

## Optimize

```{r}
theta_start1 = c(
  eta = 60,
  a = -130,   
  c = 115,
  gamma = 8000 
)

# ---- BFGS ---- 
fitBFGS1 = optim(
  par = theta_start1,
  fn = neg_loglik1,
  data = Ca2_data,
  h = 0.02,
  method = "L-BFGS-B",
  lower = c(
    eta = 1e-8,
    a = -Inf,
    c = -Inf,
    gamma = 200
  ),
  upper = c(
    eta = Inf,
    a = -1e-8,
    c = Inf,
    gamma = Inf
  )
)

fitBFGS1$value
fitBFGS1$convergence
fitBFGS1$par
```

# M2 - non-constant diffusion, full

## Likelihood

```{r}
neg_loglik2 = function(theta, data, h) {
  
  # ---- sort data ---- 
  data_old = data[-nrow(data), ]
  data_new = data[-1, ]
  n = nrow(data_old)
  
  # ---- Moments ---- 
  X_old = data_old[,1]
  Xexp  = mean(X_old)
  Xexp2 = mean(X_old^2)
  varX  = var(X_old)

  #---- theta ----
  eta   = theta[1]
  a     = theta[2]
  b     = theta[3]
  c     = theta[4]
  d     = theta[5]
  alpha = theta[6]
  beta = theta[7]
  gamma = theta[8]
  
  drift_const = 3 * a * Xexp2 + 2 * b * Xexp + c

  A = matrix(c(0, 1,
                drift_const, -eta),
              nrow = 2, byrow = TRUE)
  
  # ---- B ----
  inside_sqrt = varX + (Xexp + b / (3 * a))^2

  bxPlus = -b / (3 * a) + sqrt(inside_sqrt)
  bxMinus = -b / (3 * a) - sqrt(inside_sqrt)

  B = cbind(ifelse(X_old > 0, bxPlus, bxMinus),0)
  
  # ---- f ----
  f_step = function(df, h) {

        X = df[,1]
        V = df[,2]

        NonLin =
            a * X^3 +
            b * X^2 +
            c * X +
            d -
            drift_const * (X - B[, 1])

        out = cbind(
            X,
            V + h * NonLin
        )

        return(out)
    }

  # ---- Mu Step ----
  mu_step = function(Y, df, h) {

        expAh = expm(A * h)

        mu = as.matrix(Y - B) %*% t(expAh) + B

        return(mu)
  }

  # ---- Residual ---- 
  Y_old_half = f_step(data_old, h = 1/2*h) 
  Y_new_inv  = f_step(data_new, h = -1/2*h)
    
  Y_mu = mu_step(Y_old_half, data_old, h)
    
  Z = Y_new_inv - Y_mu
  
  # ---- Constant diffusion covariance ----
    
  first = numeric(nrow(data_old))
  det = numeric(nrow(data_old))
  
  alphaCheck = rbind(cbind(matrix(0, 2, 2), matrix(0, 2, 2)),
                     cbind(matrix(0, 2, 2), matrix(c(0,0,0,alpha), 
                                                   nrow=2, byrow=TRUE)))
  betaCheck = rbind(matrix(0, 2, 2), 
                    matrix(c(0,0,0,beta), nrow=2, byrow=TRUE))

  # ---- Kronecker ---- 
  
  I = diag(2)

  K = I %x% A + A %x% I
  
  IA = I %x% A
  AI = A %x% I
  
  Kalpha = K + alphaCheck
    
  # ---- Integrals ---- 
  
  M1 = rbind(cbind(Kalpha, alphaCheck),
               cbind(matrix(0, 4, 4), K))
  I1 = expm(M1*h)[1:4, 5:8]
  I1Bias = expm(M1*(3/2)*h)[1:4, 5:8]

  M2 = rbind(cbind(Kalpha, alphaCheck),
               cbind(matrix(0, 4, 4), IA))
  I2 = expm(M2*h)[1:4, 5:8]
  I2Bias = expm(M2*(3/2)*h)[1:4, 5:8]

  M3 = rbind(cbind(Kalpha, alphaCheck),
               cbind(matrix(0, 4, 4), AI))
  I3 = expm(M3*h)[1:4, 5:8]
  I3Bias = expm(M3*(3/2)*h)[1:4, 5:8]

  M4 = rbind(cbind(Kalpha, betaCheck),
               cbind(matrix(0, 2, 4), A)) 
  I4 = expm(M4*h)[1:4, 5:6]
  I4Bias = expm(M4*(3/2)*h)[1:4, 5:6]

  M5 = rbind( cbind(Kalpha, diag(4)),
               cbind(matrix(0, 4, 4), matrix(0, 4, 4)))

  I5 = expm(M5 * h)[1:4, 5:8]
  I5Bias = expm(M5*(3/2)*h)[1:4, 5:8]
  
  # ---- Omega ---- 
  for (i in 1:nrow(data_old)) {
    Sigma = matrix(
      c(0, 0,
        0, alpha*data_old[i,2]^2+beta*data_old[i,2]+gamma),
      nrow = 2,
      byrow = TRUE
    )
    
    xb = Y_old_half[i,]-B[i,]
    
    omegaVector = I1 %*% c(xb[1]^2, xb[1]*xb[2], xb[2]*xb[1], xb[2]^2) + 
      I2 %*% c(xb[1]*B[i,1], xb[1]*B[i,2], xb[2]*B[i,1], xb[2]*B[i,2]) + 
      I3 %*% c(B[i,1]*xb[1], B[i,1]*xb[2], B[i,2]*xb[1], B[i,2]*xb[2]) +
      I4 %*% xb +
      I5 %*% c(Sigma[1,1], Sigma[1,2], Sigma[2,1], Sigma[2,2])

    omegaVectorBias = I1Bias %*% c(xb[1]^2, xb[1]*xb[2], xb[2]*xb[1], xb[2]^2) + 
      I2Bias %*% c(xb[1]*B[i,1], xb[1]*B[i,2], xb[2]*B[i,1], xb[2]*B[i,2]) + 
      I3Bias %*% c(B[i,1]*xb[1], B[i,1]*xb[2], B[i,2]*xb[1], B[i,2]*xb[2]) +
      I4Bias %*% xb +
      I5Bias %*% c(Sigma[1,1], Sigma[1,2], Sigma[2,1], Sigma[2,2])

    # ---- Rough ----
    
    omega22Bias = as.numeric(omegaVectorBias[4])
    omega22     = as.numeric(omegaVector[4])

    det[i] = log(omega22Bias)
    first[i] = Z[i, 2]^2 / omega22
    
    # det[i] = log(omegaVectorBias[4])
    # 
    # first[i] = Z[i,2]^2 *(1/omegaVector[4])
    
    # ---- Full ----
    
    # omega = matrix(omegaVector,2,2)
    # omega_inv = solve(omega)
    # det[i] = log(det(omegaVectorBias))
    # first[i] = Z[i,1]^2 * omega_inv[1, 1] +
    # 2 * Z[i,1] * Z[i,2] * omega_inv[1, 2] +
    # Z[i,2]^2 * omega_inv[2, 2] 
  }
    
  val = sum(first) + sum((2/3)*det)
    
  return(val)
  }
```

```{r, eval = FALSE}
# ---- Starting parameters ----
  eta = 60
  a = -130   
  b = 0
  c = 115
  d = 0
  alpha = 50
  beta = 200
  gamma = 8000 
  
  # ---- sort data ---- 
  data_old = Ca2_data[-nrow(Ca2_data), ]
  data_new = Ca2_data[-1, ]
  n = nrow(data_old)
  
  # ---- Moments ---- #  
  X_old = data_old[,1]
  Xexp  = mean(X_old)
  Xexp2 = mean(X_old^2)
  varX  = var(X_old)
  
  drift_const = 3 * a * Xexp2 + 2 * b * Xexp + c

  A = matrix(c(0, 1,
                drift_const, -eta),
              nrow = 2, byrow = TRUE)
  
  # ---- B ----
  inside_sqrt = varX + (Xexp + b / (3 * a))^2

  bxPlus = -b / (3 * a) + sqrt(inside_sqrt)
  bxMinus = -b / (3 * a) - sqrt(inside_sqrt)

  B = cbind(ifelse(X_old > 0, bxPlus, bxMinus),0)
  
  # ---- f ---- 
  f_step = function(df, h) {
        
        X = df[,1]
        V = df[,2]
        
        NonLin =
            a * X^3 +
            b * X^2 +
            c * X +
            d -
            drift_const * (X -  B[, 1])
        
        out = cbind(
            X,
            V + h * NonLin
        )
  
        return(out)
    }
  
 # ---- Mu Step ----
  mu_step = function(Y, df, h) {

        expAh = expm(A * h)

        mu = as.matrix(Y - B) %*% t(expAh) + B

        return(mu)
  }

  # ---- Residual ---- 
  Y_old_half = f_step(data_old, h = 1/2*h) 
  Y_new_inv  = f_step(data_new, h = -1/2*h)
    
  Y_mu = mu_step(Y_old_half, data_old, h)
    
  Z = Y_new_inv - Y_mu
  
  # ---- Constant diffusion covariance ----
  
  alphaCheck = rbind(cbind(matrix(0, 2, 2), matrix(0, 2, 2)),
                     cbind(matrix(0, 2, 2), matrix(c(0,0,0,alpha), 
                                                   nrow=2, byrow=TRUE)))
  betaCheck = rbind(matrix(0, 2, 2), 
                    matrix(c(0,0,0,beta), nrow=2, byrow=TRUE))

  
  # ---- Kronecker ---- 
  
  I = diag(2)

  K = I %x% A + A %x% I
  
  IA = I %x% A
  AI = A %x% I
  
  Kalpha = K + alphaCheck
  # ---- Integrals ---- 
  
  M1 = rbind(cbind(Kalpha, alphaCheck),
               cbind(matrix(0, 4, 4), K))
  I1 = expm(M1*h)[1:4, 5:8]
  I1Bias = expm(M1*(3/2)*h)[1:4, 5:8]

  M2 = rbind(cbind(Kalpha, alphaCheck),
               cbind(matrix(0, 4, 4), IA))
  I2 = expm(M2*h)[1:4, 5:8]
  I2Bias = expm(M2*(3/2)*h)[1:4, 5:8]

  M3 = rbind(cbind(Kalpha, alphaCheck),
               cbind(matrix(0, 4, 4), AI))
  I3 = expm(M3*h)[1:4, 5:8]
  I3Bias = expm(M3*(3/2)*h)[1:4, 5:8]

  M4 = rbind(cbind(Kalpha, betaCheck),
               cbind(matrix(0, 2, 4), A)) 
  I4 = expm(M4*h)[1:4, 5:6]
  I4Bias = expm(M4*(3/2)*h)[1:4, 5:6]

  M5 = rbind( cbind(Kalpha, diag(4)),
               cbind(matrix(0, 4, 4), matrix(0, 4, 4)))

  I5 = expm(M5 * h)[1:4, 5:8]
  I5Bias = expm(M5*(3/2)*h)[1:4, 5:8]

  # ---- Omega ---- 
      
  first = numeric(nrow(data_old))
  det = numeric(nrow(data_old))
  
  for (i in 1:nrow(data_old)) {
    Sigma = matrix(
      c(0, 0,
        0, alpha*data_old[i,2]^2+beta*data_old[i,2]+gamma),
      nrow = 2,
      byrow = TRUE
    )
    
    xb = Y_old_half[i,]-B[i,]
    
    omegaVector = I1 %*% as.vector(xb %*% t(xb)) +
      I2 %*% as.vector(xb%*% t(B[i,])) +
      I3 %*% as.vector(B[i,]%*%t(xb)) +
      I4 %*% as.vector(xb) +
      I5 %*% as.vector(Sigma)

    omegaVectorBias = I1Bias %*% as.vector(xb %*% t(xb)) +
      I2Bias %*% as.vector(xb%*% t(B[i,])) +
      I3Bias %*% as.vector(B[i,]%*%t(xb)) +
      I4Bias %*% as.vector(xb) +
      I5Bias %*% as.vector(Sigma)

    # ---- Rough ----
    det[i] = log(omegaVectorBias[4])
  
    first[i] = Z[i,2]^2 *(1/omegaVector[4])
    
    # ---- Full ----
    
    # omega = matrix(omegaVector,2,2)
    # omega_inv = solve(omega)
    # det[i] = log(det(omegaVectorBias))
    # first[i] = Z[i,1]^2 * omega_inv[1, 1] +
    # 2 * Z[i,1] * Z[i,2] * omega_inv[1, 2] +
    # Z[i,2]^2 * omega_inv[2, 2] 
  }
    
  val = sum(first) + sum((2/3)*det)
```

## Optimize

```{r}
# ---- Starting parameter ----
theta_start2 = c(
  eta = 60,
  a = -130,   
  b = 0,
  c = 115,
  d = 0,
  alpha = 50,
  beta = 200,
  gamma = 8000 
)

# ---- Optimizer -----
fitBFGS2 = optim(
  par = theta_start2,
  fn = neg_loglik2,
  data = Ca2_data,
  h = 0.02,
  method = "L-BFGS-B",
  lower = c(
    eta   = 1,
    a     = -Inf,
    b     = -Inf,
    c     = -Inf,
    d     = -Inf,
    alpha = 1e-8,
    beta  = -Inf,
    gamma = 1e-8), 
  upper = c(
    eta   = Inf,
    a     = -1e-8,
    b     = Inf,
    c     = Inf,
    d     = Inf,
    alpha = Inf,
    beta  = Inf,
    gamma = Inf)
)

# ---- Fit ---- 
fitBFGS2$par
fitBFGS2$value
fitBFGS2$convergence
```

# M3 - non-constant diffusion, b=d=0

## Likelihood

```{r}
neg_loglik3 = function(theta, data, h) {
  
  # ---- sort data ---- 
  data_old = data[-nrow(data), ]
  data_new = data[-1, ]
  n = nrow(data_old)
  
  # ---- Moments ---- 
  X_old = data_old[,1]
  Xexp  = mean(X_old)
  Xexp2 = mean(X_old^2)
  varX  = var(X_old)

  #---- theta ----
  eta   = theta[1]
  a     = theta[2]
  b     = 0
  c     = theta[3]
  d     = 0
  alpha = theta[4]
  beta = theta[5]
  gamma = theta[6]
  
  drift_const = 3 * a * Xexp2 + 2 * b * Xexp + c

  A = matrix(c(0, 1,
                drift_const, -eta),
              nrow = 2, byrow = TRUE)
  
  # ---- B ----
  inside_sqrt = varX + (Xexp + b / (3 * a))^2

  bxPlus = -b / (3 * a) + sqrt(inside_sqrt)
  bxMinus = -b / (3 * a) - sqrt(inside_sqrt)

  B = cbind(ifelse(X_old > 0, bxPlus, bxMinus),0)
  
  # ---- f ----
  f_step = function(df, h) {

        X = df[,1]
        V = df[,2]

        NonLin =
            a * X^3 +
            b * X^2 +
            c * X +
            d -
            drift_const * (X - B[, 1])

        out = cbind(
            X,
            V + h * NonLin
        )

        return(out)
    }

  # ---- Mu Step ----
  mu_step = function(Y, df, h) {

        expAh = expm(A * h)

        mu = as.matrix(Y - B) %*% t(expAh) + B

        return(mu)
  }

  # ---- Residual ---- 
  Y_old_half = f_step(data_old, h = 1/2*h) 
  Y_new_inv  = f_step(data_new, h = -1/2*h)
    
  Y_mu = mu_step(Y_old_half, data_old, h)
    
  Z = Y_new_inv - Y_mu
  
  # ---- Constant diffusion covariance ----
    
  first = numeric(nrow(data_old))
  det = numeric(nrow(data_old))
  
  alphaCheck = rbind(cbind(matrix(0, 2, 2), matrix(0, 2, 2)),
                     cbind(matrix(0, 2, 2), matrix(c(0,0,0,alpha), 
                                                   nrow=2, byrow=TRUE)))
  betaCheck = rbind(matrix(0, 2, 2), 
                    matrix(c(0,0,0,beta), nrow=2, byrow=TRUE))

  # ---- Kronecker ---- 
  
  I = diag(2)

  K = I %x% A + A %x% I
  
  IA = I %x% A
  AI = A %x% I
  
  Kalpha = K + alphaCheck
    
  # ---- Integrals ---- 
  
  M1 = rbind(cbind(Kalpha, alphaCheck),
               cbind(matrix(0, 4, 4), K))
  I1 = expm(M1*h)[1:4, 5:8]
  I1Bias = expm(M1*(3/2)*h)[1:4, 5:8]

  M2 = rbind(cbind(Kalpha, alphaCheck),
               cbind(matrix(0, 4, 4), IA))
  I2 = expm(M2*h)[1:4, 5:8]
  I2Bias = expm(M2*(3/2)*h)[1:4, 5:8]

  M3 = rbind(cbind(Kalpha, alphaCheck),
               cbind(matrix(0, 4, 4), AI))
  I3 = expm(M3*h)[1:4, 5:8]
  I3Bias = expm(M3*(3/2)*h)[1:4, 5:8]

  M4 = rbind(cbind(Kalpha, betaCheck),
               cbind(matrix(0, 2, 4), A)) 
  I4 = expm(M4*h)[1:4, 5:6]
  I4Bias = expm(M4*(3/2)*h)[1:4, 5:6]

  M5 = rbind( cbind(Kalpha, diag(4)),
               cbind(matrix(0, 4, 4), matrix(0, 4, 4)))

  I5 = expm(M5 * h)[1:4, 5:8]
  I5Bias = expm(M5*(3/2)*h)[1:4, 5:8]
  
  # ---- Omega ---- 
  for (i in 1:nrow(data_old)) {
    Sigma = matrix(
      c(0, 0,
        0, alpha*data_old[i,2]^2+beta*data_old[i,2]+gamma),
      nrow = 2,
      byrow = TRUE
    )
    
    xb = Y_old_half[i,]-B[i,]
    
    omegaVector = I1 %*% c(xb[1]^2, xb[1]*xb[2], xb[2]*xb[1], xb[2]^2) + 
      I2 %*% c(xb[1]*B[i,1], xb[1]*B[i,2], xb[2]*B[i,1], xb[2]*B[i,2]) + 
      I3 %*% c(B[i,1]*xb[1], B[i,1]*xb[2], B[i,2]*xb[1], B[i,2]*xb[2]) +
      I4 %*% xb +
      I5 %*% c(Sigma[1,1], Sigma[1,2], Sigma[2,1], Sigma[2,2])

    omegaVectorBias = I1Bias %*% c(xb[1]^2, xb[1]*xb[2], xb[2]*xb[1], xb[2]^2) + 
      I2Bias %*% c(xb[1]*B[i,1], xb[1]*B[i,2], xb[2]*B[i,1], xb[2]*B[i,2]) + 
      I3Bias %*% c(B[i,1]*xb[1], B[i,1]*xb[2], B[i,2]*xb[1], B[i,2]*xb[2]) +
      I4Bias %*% xb +
      I5Bias %*% c(Sigma[1,1], Sigma[1,2], Sigma[2,1], Sigma[2,2])

    # ---- Rough ----
    
    omega22Bias = as.numeric(omegaVectorBias[4])
    omega22     = as.numeric(omegaVector[4])

    det[i] = log(omega22Bias)
    first[i] = Z[i, 2]^2 / omega22
    
    # det[i] = log(omegaVectorBias[4])
    # 
    # first[i] = Z[i,2]^2 *(1/omegaVector[4])
    
    # ---- Full ----
    
    # omega = matrix(omegaVector,2,2)
    # omega_inv = solve(omega)
    # det[i] = log(det(omegaVectorBias))
    # first[i] = Z[i,1]^2 * omega_inv[1, 1] +
    # 2 * Z[i,1] * Z[i,2] * omega_inv[1, 2] +
    # Z[i,2]^2 * omega_inv[2, 2] 
  }
    
  val = sum(first) + sum((2/3)*det)
    
  return(val)
  }
```

## Optimize

```{r}
# ---- Starting parameter ----
theta_start3 = c(
  eta = 60,
  a = -130,  
  c = 115,
  alpha = 50,
  beta = 200,
  gamma = 8000 
)

# ---- Optimizer -----
fitBFGS3 = optim(
  par = theta_start3,
  fn = neg_loglik3,
  data = Ca2_data,
  h = 0.02,
  method = "L-BFGS-B",
  lower = c(
    eta   = 1,
    a     = -Inf,
    c     = -Inf,
    alpha = 1e-8,
    beta  = -Inf,
    gamma = 1e-8), 
  upper = c(
    eta   = Inf,
    a     = -1e-8,
    c     = Inf,
    alpha = Inf,
    beta  = Inf,
    gamma = Inf)
)

# ---- Fit ---- 
fitBFGS3$par
fitBFGS3$value
fitBFGS3$convergence
```

# Simulated Paths

## M0 for simulated paths

```{r}
#theta = fitNM$par
theta = fitBFGS$par

eta   = theta["eta"]
a     = theta["a"]
b     = theta["b"]
c     = theta["c"]
d     = theta["d"]
gamma = theta["gamma"]

set.seed(16)

h = 0.0001
n = 500000

SimValues = data.frame(age = numeric(n),
                     X = numeric(n), 
                     V = numeric(n))
SimValues$age[1]=80.02
SimValues$X[1]=1.11
SimValues$V[1]=5.43


for (i in 2:n){
  SimValues$age[i]=SimValues$age[i - 1]-h
  X_prev = SimValues$X[i - 1]
  V_prev = SimValues$V[i - 1]
  SimValues$X[i]=X_prev + h*V_prev
  SimValues$V[i]=V_prev+
    h*(-eta*V_prev+a*X_prev^3+b*X_prev^2+c*X_prev+d)+sqrt(h)*sqrt(gamma)*rnorm(1)
}

SimData = SimValues[seq(1, nrow(SimValues), by = 200), ]

```

### Plots

```{r}
# ---- Plot of simulated data ---- (ALSO ADD DENSITY PLOT)
SimPlot1 = ggplot(SimData, aes(x = age, y = X)) +
    geom_line(linewidth = 0.7) +
    labs(
        title = "Simulated X",
        x = "Age (kiloyears before present)",
        y = expression(X[t])
    ) +
    theme_minimal(base_size = 13) +
    theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        panel.grid.minor = element_blank()
    ) +
    scale_x_reverse()

SimPlot2 = ggplot(SimData, aes(x = age, y = V)) +
    geom_line(linewidth = 0.7) +
    labs(
        title = "Simulated V",
        x = "Age (kiloyears before present)",
        y = expression(V[t])
    ) +
    theme_minimal(base_size = 13) +
    theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        panel.grid.minor = element_blank()
    ) +
    scale_x_reverse()
grid.arrange(SimPlot1,SimPlot2, ncol = 1)
```

```{r}
# ---- Simulated paths on real data ----
RealSimX = ggplot(Ca2_final, aes(x = age, y = X)) +
  geom_line(linewidth = 0.7) +
  labs(
    title = expression("Ice Land Core Greenland: X[t] =-log(Ca"^{"2+"}*"))"),
    x = "Age (kiloyears before present)",
    y = expression(X[t])
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank() 
  ) +
  scale_x_reverse() +
  geom_line(
    data = SimData,
    aes(x = age, y = X),
    linewidth = 0.7,
    linetype = "dashed", 
    color = 'darkred'
  )

RealSimV =  ggplot(Ca2_final, aes(x = age, y = V)) +
  geom_line(linewidth = 0.7) +
  labs(
    title = "Estimated Velocity Process",
    x = "Age (kiloyears before present)",
    y = expression(V[t])
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  scale_x_reverse() + 
  geom_line(
    data = SimData,
    aes(x = age, y = V),
    linewidth = 0.7,
    linetype = "dashed", 
    color = 'darkred'
  )

grid.arrange(RealSimX,RealSimV, ncol = 1)
```

```{r}
DensitySimX = ggplot(Ca2_final, aes(x =X)) +
  geom_density(linewidth = 0.7) +
  labs(
    title = "Density",
    x = "Age (kiloyears before present)",
    y = expression(V[t])
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  scale_x_reverse() +
  geom_density(
    data = SimData,
    aes(x = X),
    linewidth = 0.7,
    linetype = "dashed", 
    color = 'darkred'
  ) 

DensitySimV = ggplot(Ca2_final, aes(x =V)) +
  geom_density(linewidth = 0.7) +
  labs(
    title = "Density",
    x = "Age (kiloyears before present)",
    y = expression(V[t])
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  scale_x_reverse() +
  geom_density(
    data = SimData,
    aes(x = V),
    linewidth = 0.7,
    linetype = "dashed", 
    color = 'darkred'
  )

grid.arrange(DensitySimX,DensitySimV, ncol = 1)
```

```{r}
p_dens_X = ggplot() +
  geom_density(
    data = SimMany,
    aes(y = X, x = after_stat(density), group = sim_id),
    color = "firebrick",
    alpha = 0.10,
    linewidth = 0.4
  ) +
  geom_density(
    data = Ca2_final,
    aes(y = X, x = after_stat(density)),
    color = "black",
    linewidth = 1
  ) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r}

base_theme = theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  )

# ---- X_t over time ----
RealSimX = ggplot(Ca2_final, aes(x = age, y = X)) +
  geom_line(linewidth = 0.7) +
  geom_line(
    data = SimData2,
    aes(x = age, y = X),
    linewidth = 0.7,
    linetype = "dashed",
    colour = "darkred"
  ) +
  labs(
    title = expression(X[t]),
    x = NULL,
    y = expression(X[t])
  ) +
  scale_x_reverse() +
  base_theme +
  theme(
    axis.title.x = element_blank(),
    plot.margin = margin(5.5, 5.5, 2, 5.5)
  )

# ---- V_t over time ----
RealSimV = ggplot(Ca2_final, aes(x = age, y = V)) +
  geom_line(linewidth = 0.7) +
  geom_line(
    data = SimData2,
    aes(x = age, y = V),
    linewidth = 0.7,
    linetype = "dashed",
    colour = "darkred"
  ) +
  labs(
    title = expression(V[t]),
    x = "Age (kiloyears before present)",
    y = expression(V[t])
  ) +
  scale_x_reverse() +
  base_theme +
  theme(
    plot.margin = margin(2, 5.5, 5.5, 5.5)
  )

# ---- Density of X_t ----
DensitySimX = ggplot() +
  geom_density(
    data = Ca2_final,
    aes(x = X),
    linewidth = 0.7,
    colour = "black"
  ) +
  geom_density(
    data = SimData2,
    aes(x = X),
    linewidth = 0.7,
    linetype = "dashed",
    colour = "darkred",
    fill = "darkred",
    alpha = 0.25
  ) +
  coord_flip() +
  labs(
    title = "Density",
    x = NULL,
    y = NULL
  ) +
  base_theme +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    plot.margin = margin(5.5, 5.5, 2, 2)
  )

# ---- Density of V_t ----
DensitySimV = ggplot() +
  geom_density(
    data = Ca2_final,
    aes(x = V),
    linewidth = 0.7,
    colour = "black"
  ) +
  geom_density(
    data = SimData2,
    aes(x = V),
    linewidth = 0.7,
    linetype = "dashed",
    colour = "darkred",
    fill = "darkred",
    alpha = 0.25
  ) +
  coord_flip() +
  labs(
    title = NULL,
    x = NULL,
    y = NULL
  ) +
  base_theme +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    plot.margin = margin(2, 5.5, 5.5, 2)
  )

# ---- Combine ----
((RealSimX | DensitySimX) / (RealSimV | DensitySimV)) +
  plot_layout(widths = c(2, 1), heights = c(1, 1)) +
  plot_annotation(
    title = "Greenland Ice-Core Data and Simulated Paths for Model 2: Non-constant Diffusion",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
  )
```

```{r}
# ----Refit ----
  fit_bootSIM = optim(
    par = theta_start,
    fn = neg_loglik,
    data = SimData,
    h = 0.02,
    method = "L-BFGS-B",
    lower = c(
        eta = 1,
        a = -Inf,
        b = -Inf,
        c = -Inf,
        d = -Inf,
        gamma = 1e-8),
  upper = c(
        eta = Inf,
        a = -1e-8,
        b = Inf,
        c = Inf,
        d = Inf,
        gamma = Inf)
  )
```

## M2 simulated path

```{r}
theta2 = fitBFGS2$par

eta   = theta2["eta"]
a     = theta2["a"]
b     = theta2["b"]
c     = theta2["c"]
d     = theta2["d"]
alpha = theta2['alpha']
beta = theta2['beta']
gamma = theta["gamma"]

set.seed(16)

h = 0.0001
n = 500000

SimValues2 = data.frame(age = numeric(n),
                     X = numeric(n), 
                     V = numeric(n))
SimValues2$age[1]=80.02
SimValues2$X[1]=1.11
SimValues2$V[1]=5.43


for (i in 2:n){
  SimValues2$age[i]=SimValues2$age[i - 1]-h
  X_prev = SimValues2$X[i - 1]
  V_prev = SimValues2$V[i - 1]
  SimValues2$X[i]=X_prev + h*V_prev
  SimValues2$V[i]=V_prev+
    h*(-eta*V_prev+a*X_prev^3+b*X_prev^2+c*X_prev+d)+sqrt(h)*sqrt(alpha*V_prev^2+beta*V_prev+gamma)*rnorm(1)
}

SimData2 = SimValues2[seq(1, nrow(SimValues2), by = 200), ]
```

### Plots

```{r}
# ---- Plot of simulated data ---- (ALSO ADD DENSITY PLOT)
SimPlotX2 = ggplot(SimData2, aes(x = age, y = X)) +
    geom_line(linewidth = 0.7) +
    labs(
        title = "Simulated X",
        x = "Age (kiloyears before present)",
        y = expression(X[t])
    ) +
    theme_minimal(base_size = 13) +
    theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        panel.grid.minor = element_blank()
    ) +
    scale_x_reverse()

SimPlotV2 = ggplot(SimData2, aes(x = age, y = V)) +
    geom_line(linewidth = 0.7) +
    labs(
        title = "Simulated V",
        x = "Age (kiloyears before present)",
        y = expression(V[t])
    ) +
    theme_minimal(base_size = 13) +
    theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        panel.grid.minor = element_blank()
    ) +
    scale_x_reverse()
grid.arrange(SimPlotX2,SimPlotV2, ncol = 1)
```

```{r}
# ---- Simulated paths on real data ----
RealSimX2 = ggplot(Ca2_final, aes(x = age, y = X)) +
  geom_line(linewidth = 0.7) +
  labs(
    title = expression("Ice Land Core Greenland: X[t] =-log(Ca"^{"2+"}*"))"),
    x = "Age (kiloyears before present)",
    y = expression(X[t])
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank() 
  ) +
  scale_x_reverse() +
  geom_line(
    data = SimData2,
    aes(x = age, y = X),
    linewidth = 0.7,
    linetype = "dashed", 
    color = 'darkred'
  )

RealSimV2 =  ggplot(Ca2_final, aes(x = age, y = V)) +
  geom_line(linewidth = 0.7) +
  labs(
    title = "Estimated Velocity Process",
    x = "Age (kiloyears before present)",
    y = expression(V[t])
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  scale_x_reverse() + 
  geom_line(
    data = SimData2,
    aes(x = age, y = V),
    linewidth = 0.7,
    linetype = "dashed", 
    color = 'darkred'
  )

grid.arrange(RealSimX2,RealSimV2, ncol = 1)
```

```{r}
# ---- Reorder for overview + plots ----
SimParameters_clean = BootParameters0[complete.cases(BootParameters0), 1:6 ]

boot_se = apply(SimParameters_clean, 2, sd)
boot_mean = apply(SimParameters_clean, 2, mean)

boot_ci = apply(
  SimParameters_clean,
  2,
  quantile,
  probs = c(0.025, 0.975)
)

boot_bias = colMeans(SimParameters_clean) - theta_hat

# ---- Plotting simulation (Bootstrap) results ----
SimParametersClean_long = SimParameters_clean %>% 
  pivot_longer(
    cols = everything(),
    names_to = "parameter",
    values_to = "estimate"
  )

fit_values = tibble(
  parameter = names(fitBFGS$par),
  fit = as.numeric(fitBFGS$par)
)

# Density plot with fitted values as vertical lines
ggplot(SimParametersClean_long, aes(x = estimate)) +
  geom_density(fill = "gray80", color = "black") +
  geom_vline(data = fit_values, aes(xintercept = fit), color = "red", linetype = "dashed", size = 1) +
  facet_wrap(~ parameter, scales = "free", ncol = 2) +
  theme_minimal() +
  labs(title = "Density of Simulated Parameters with MLEs", x = "Estimate", y = "Density")

# Histogram plot with fitted values as vertical lines
ggplot(SimParametersClean_long, aes(x = estimate)) +
  geom_histogram(bins = 5, fill = "gray80", color = "black") +
  geom_vline(data = fit_values, aes(xintercept = fit), color = "red", linetype = "dashed", size = 1) +
  facet_wrap(~ parameter, scales = "free", ncol = 2) +
  theme_minimal() +
  labs(title = "Histogram of Simulated Parameters with MLEs", x = "Estimate", y = "Count")
```

# Bootstrap Simulations

```{r}

# ---- Simulate Model 0 ----
SimulateModel0_sub = function(theta, h_sim, n_data, subselect) {
  
  eta   = theta["eta"]
  a     = theta["a"]
  b     = theta["b"]
  c     = theta["c"]
  d     = theta["d"]
  gamma = theta["gamma"]
  
  X = numeric(n_data)
  V = numeric(n_data)
  
  # Initial condition
  X_curr = 1.11
  V_curr = 5.43
  
  X[1] = X_curr
  V[1] = V_curr
  
  save_idx = 2
  
  for (i in 2:((n_data - 1) * subselect + 1)) {
    
    X_prev = X_curr
    V_prev = V_curr
    
    X_curr = X_prev + h_sim * V_prev
    
    V_curr = V_prev +
      h_sim * (-eta * V_prev +
                 a * X_prev^3 +
                 b * X_prev^2 +
                 c * X_prev +
                 d) +
      sqrt(h_sim) * sqrt(gamma) * rnorm(1)
    
    if ((i - 1) %% subselect == 0) {
      X[save_idx] = X_curr
      V[save_idx] = V_curr
      save_idx = save_idx + 1
    }
  }
  
  data.frame(X = X, V = V)
}

# ---- Simulate Model 1 ----
SimulateModel1_sub = function(theta, h_sim, n_data, subselect) {
  
  eta   = theta["eta"]
  a     = theta["a"]
  b     = 0 
  c     = theta["c"]
  d     = 0
  gamma = theta["gamma"]
  
  X = numeric(n_data)
  V = numeric(n_data)
  
  # Initial condition
  X_curr = 1.11
  V_curr = 5.43
  
  X[1] = X_curr
  V[1] = V_curr
  
  save_idx = 2
  
  for (i in 2:((n_data - 1) * subselect + 1)) {
    
    X_prev = X_curr
    V_prev = V_curr
    
    X_curr = X_prev + h_sim * V_prev
    
    V_curr = V_prev +
      h_sim * (-eta * V_prev +
                 a * X_prev^3 +
                 b * X_prev^2 +
                 c * X_prev +
                 d) +
      sqrt(h_sim) * sqrt(gamma) * rnorm(1)
    
    if ((i - 1) %% subselect == 0) {
      X[save_idx] = X_curr
      V[save_idx] = V_curr
      save_idx = save_idx + 1
    }
  }
  
  data.frame(X = X, V = V)
}
```

```{r}
#theta_hat = fitNM$par
theta_hat = fitBFGS$par

theta_start = c(
  eta   = 60,
  a     = -130,
  b     = 0,
  c     = 115,
  d     = 0,
  gamma = 8000
)

lower = c(
  eta = 1,
  a = -Inf,
  b = -Inf,
  c = -Inf,
  d = -Inf,
  gamma = 1e-8
)

upper = c(
  eta = Inf,
  a = -1e-8,
  b = Inf,
  c = Inf,
  d = Inf,
  gamma = Inf
)

h_sim = 0.0001
subselect = 200
M = 500
n = 500000
h_fit = h_sim * subselect

BootParameters= data.frame(
  eta = rep(NA, M),
  a = rep(NA, M), 
  b = rep(NA, M), 
  c = rep(NA, M),
  d = rep(NA, M),
  gamma = rep(NA, M),
  convergence = rep(NA, M),
  value = rep(NA,M)
)


for (m in 1:M) {
  
  # ---- Simulate ---- 

  SimData_sub = SimulateModel0_sub(theta = theta_hat, h_sim = h_sim, n_data = 2499, subselect = 200)
  
  print(2)
  # ----Refit ----
  fit_boot0 = optim(
    par = theta_hat,
    fn = neg_loglik,
    data = SimData_sub,
    h = h_fit,
    method = "L-BFGS-B",
    lower = lower,
  upper = upper,
  )

  # ---- Store estimates ---- 
  BootParameters[m, ] = c(fit_boot0$par, fit_boot0$convergence, fit_boot0$value)
  
  print(paste("Boot: ",m, fit_boot0$par))
}


```

```{r}
# ---- Reorder for overview + plots ----
SimParameters_clean = BootParameters[complete.cases(BootParameters), 1:6 ]

boot_se = apply(SimParameters_clean, 2, sd)
boot_mean = apply(SimParameters_clean, 2, mean)

boot_ci = apply(
  SimParameters_clean,
  2,
  quantile,
  probs = c(0.025, 0.975)
)

boot_bias = colMeans(SimParameters_clean) - theta_hat

# ---- Plotting simulation (Bootstrap) results ----
SimParametersClean_long = SimParameters_clean %>% 
  pivot_longer(
    cols = everything(),
    names_to = "parameter",
    values_to = "estimate"
  )

fit_values = tibble(
  parameter = names(fitBFGS$par),
  fit = as.numeric(fitBFGS$par)
)

# Density plot with fitted values as vertical lines
ggplot(SimParametersClean_long, aes(x = estimate)) +
  geom_density(fill = "gray80", color = "black") +
  geom_vline(data = fit_values, aes(xintercept = fit), color = "red", linetype = "dashed", size = 1) +
  facet_wrap(~ parameter, scales = "free", ncol = 2) +
  theme_minimal() +
  labs(title = "Density of Simulated Parameters with MLEs", x = "Estimate", y = "Density")

# Histogram plot with fitted values as vertical lines
ggplot(SimParametersClean_long, aes(x = estimate)) +
  geom_histogram(bins = 5, fill = "gray80", color = "black") +
  geom_vline(data = fit_values, aes(xintercept = fit), color = "red", linetype = "dashed", size = 1) +
  facet_wrap(~ parameter, scales = "free", ncol = 2) +
  theme_minimal() +
  labs(title = "Histogram of Simulated Parameters with MLEs", x = "Estimate", y = "Count")
```

```{r}
param_labs = c(
  eta   = expression(eta),
  a     = "a",
  b     = "b",
  c     = "c",
  d     = "d",
  gamma = expression(gamma)
)

# Density plot
p_density = ggplot(SimParametersClean_long, aes(x = estimate)) +
  geom_density(
    fill = "grey85",
    color = "grey25",
    linewidth = 0.6,
    alpha = 0.8
  ) +
  geom_vline(
    data = fit_values,
    aes(xintercept = fit),
    color = "red3",
    linetype = "dashed",
    linewidth = 0.8
  ) +
  facet_wrap(
    ~ parameter,
    scales = "free",
    ncol = 2,
    labeller = as_labeller(param_labs, default = label_parsed)
  ) +
  labs(
    title = "Simulation distributions of parameter estimates with true parameter estimated by Model 0",
    subtitle = "Dashed red lines show fitted MLEs Model 0 from the observed data",
    x = "Estimate",
    y = "Density"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey35"),
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

p_hist = ggplot(SimParametersClean_long, aes(x = estimate)) +
  geom_histogram(
    bins = 15,
    fill = "grey85",
    color = "grey25",
    linewidth = 0.3
  ) +
  geom_vline(
    data = fit_values,
    aes(xintercept = fit),
    color = "red3",
    linetype = "dashed",
    linewidth = 0.8
  ) +
  facet_wrap(
    ~ parameter,
    scales = "free",
    ncol = 2,
    labeller = as_labeller(param_labs, default = label_parsed)
  ) +
  labs(
    title = "Simulation histograms of parameter estimates with true parameter estimated by Model 0",
    subtitle = "Dashed red lines show fitted MLEs Model 0 from the observed data",
    x = "Estimate",
    y = "Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey35"),
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

p_density
```

## M0 : Full

```{r}

h_sim = 0.0001
subselect = 200
M = 500
n = 500000
h_fit = h_sim * subselect

theta_hat_full = fitBFGS_full$par

BootParametersFull1= data.frame(
  eta = rep(NA, M),
  a = rep(NA, M), 
  b = rep(NA, M), 
  c = rep(NA, M),
  d = rep(NA, M),
  gamma = rep(NA, M),
  convergence = rep(NA, M),
  value = rep(NA,M)
)


for (m in 1:M) {
  
  # ---- Simulate ---- 

  SimData_sub = SimulateModel0_sub(theta = theta_hat_full, h_sim = h_sim, n_data = 2499, subselect = 200)
  
  print(2)
  # ----Refit ----
  fit_boot0_full = optim(
    par = theta_hat_full,
    fn = neg_loglik_full,
    data = SimData_sub,
    h = h_fit,
    method = "L-BFGS-B",
    lower = lower,
  upper = upper,
  )

  # ---- Store estimates ---- 
  BootParametersFull1[m, ] = c(fit_boot0_full$par, fit_boot0_full$convergence, fit_boot0_full$value)
  
  print(paste("Boot: ",m, fit_boot0_full$par))
}


# ---- Uncorrected ----

theta_hat_no_bias = fitBFGS_no_bias$par

BootParametersNoBias1= data.frame(
  eta = rep(NA, M),
  a = rep(NA, M), 
  b = rep(NA, M), 
  c = rep(NA, M),
  d = rep(NA, M),
  gamma = rep(NA, M),
  convergence = rep(NA, M),
  value = rep(NA,M)
)


for (m in 1:M) {
  
  # ---- Simulate ---- 

  SimData_sub = SimulateModel0_sub(theta = theta_hat_no_bias, h_sim = h_sim, n_data = 2499, subselect = 200)
  
  print(2)
  # ----Refit ----
  fit_boot0_no_bias = optim(
    par = theta_start0,
    fn = neg_loglik_no_bias,
    data = SimData_sub,
    h = h_fit,
    method = "L-BFGS-B",
    lower = lower,
  upper = upper,
  )

  # ---- Store estimates ---- 
  BootParametersNoBias1[m, ] = c(fit_boot0_no_bias$par, fit_boot0_no_bias$convergence, fit_boot0_no_bias$value)
  
  print(paste("Boot: ",m, fit_boot0_no_bias$par, fit_boot0_no_bias$value))
}
```

```{r}
# ---- Bias ----
bias_rough = colMeans(BootParameters[, 1:6], na.rm = TRUE) -
  fitBFGS$par[1:6]

bias_full = colMeans(BootParametersFull[, 1:6], na.rm = TRUE) -
  fitBFGS_full$par[1:6]

bias_uncorrected = colMeans(BootParametersNoBias[, 1:6], na.rm = TRUE) -
  fitBFGS_no_bias$par[1:6]

bias_rough
bias_full
bias_uncorrected
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# ---- keep only parameter columns ----
param_names = c("eta", "a", "b", "c", "d", "gamma")

# ---- bootstrap draws in long format ----
boot_long = bind_rows(
  BootParameters %>%
    select(all_of(param_names)) %>%
    mutate(Model = "Rough"),
  
  BootParametersFull %>%
    select(all_of(param_names)) %>%
    mutate(Model = "Full"),
  
  BootParametersNoBias %>%
    select(all_of(param_names)) %>%
    mutate(Model = "Uncorrected")
) %>%
  pivot_longer(
    cols = all_of(param_names),
    names_to = "Parameter",
    values_to = "Estimate"
  )

# ---- vertical line values in long format ----
true_vals = bind_rows(
  data.frame(
    Model = "Rough",
    Parameter = names(fitBFGS$par),
    TrueValue = as.numeric(fitBFGS$par)
  ),
  data.frame(
    Model = "Full",
    Parameter = names(fitBFGS_full$par),
    TrueValue = as.numeric(fitBFGS_full$par)
  ),
  data.frame(
    Model = "Uncorrected",
    Parameter = names(fitBFGS_no_bias$par),
    TrueValue = as.numeric(fitBFGS_no_bias$par)
  )
)

# ---- make sure model order is nice ----
boot_long$Model = factor(boot_long$Model, levels = c("Rough", "Full", "Uncorrected"))
true_vals$Model = factor(true_vals$Model, levels = c("Rough", "Full", "Uncorrected"))

# ---- plot ----
model_cols = c(
  "Rough" = "darkred",
  "Full" = "darkgreen",
  "Uncorrected" = "darkblue"
)

param_labs = c(
  eta = expression(eta),
  a = "a",
  b = "b",
  c = "c",
  d = "d",
  gamma = expression(gamma)
)


combinedplot = ggplot(boot_long, aes(x = Estimate, color = Model, fill = Model)) +
  geom_density(alpha = 0.2, linewidth = 1) +
  geom_vline(
    data = true_vals,
    aes(xintercept = TrueValue, color = Model),
    linetype = "dashed",
    linewidth = 0.9,
    show.legend = FALSE
  ) +
  facet_wrap(~ Parameter, scales = "free", ncol = 2, labeller = as_labeller(param_labs, default = label_parsed)) +
  labs(
    title = "Bootstrap densities by parameter",
    x = "Parameter estimate",
    y = "Density"
  ) +
  scale_color_manual(values = model_cols) +
  scale_fill_manual(values = model_cols) +
  theme_minimal(base_size = 13) +
 theme(
  plot.title = element_text(hjust = 0.5, face = "bold"),
  legend.position = "top",
  legend.title = element_blank()
)
```

```{r}
theta_hat1 = fitBFGS1$par

theta_start1 = c(
  eta   = 60,
  a     = -130,
  c     = 115,
  gamma = 8000
)

lower1 = c(
  eta = 1,
  a = -Inf,
  c = -Inf,
  gamma = 1e-8
)

upper1 = c(
  eta = Inf,
  a = -1e-8,
  c = Inf,
  gamma = Inf
)

h_sim = 0.0001
subselect = 200
M = 500
n = 500000
h_fit = h_sim * subselect

BootParameters1 = data.frame(
  eta = rep(NA, M),
  a = rep(NA, M), 
  c = rep(NA, M),
  gamma = rep(NA, M),
  convergence = rep(NA, M),
  value = rep(NA,M)
)


for (m in 1:M) {
  
  # ---- Simulate ---- 
  
  SimData_sub1 = SimulateModel1_sub(theta = theta_hat, h_sim = h_sim, n_data = 2499, subselect = 200)

  # ----Refit ----
  fit_boot1 = optim(
    par = theta_hat1,
    fn = neg_loglik1,
    data = SimData_sub1,
    h = 0.02,
    method = "L-BFGS-B",
    lower = lower1,
  upper = upper1,
  )
  # ---- Store estimates ---- 
  BootParameters1[m, ] = c(fit_boot1$par, fit_boot1$convergence, fit_boot1$value)
  
  print(paste("Boot: ",m, fit_boot1$par))
}

```

```{r}
# ---- Reorder for overview + plots ----
SimParameters_clean1 = BootParameters1[complete.cases(BootParameters1), 1:4 ]

boot_se1 = apply(SimParameters_clean1, 2, sd)
boot_mean1 = apply(SimParameters_clean1, 2, mean)

boot_ci1 = apply(
  SimParameters_clean1,
  2,
  quantile,
  probs = c(0.025, 0.975)
)

boot_bias1 = colMeans(SimParameters_clean1) - theta_hat1

# ---- Plotting simulation (Bootstrap) results ----
SimParametersClean_long1 = SimParameters_clean1 %>% 
  pivot_longer(
    cols = everything(),
    names_to = "parameter",
    values_to = "estimate"
  )

fit_values1 = tibble(
  parameter = names(fitBFGS1$par),
  fit = as.numeric(fitBFGS1$par)
)

# Density plot with fitted values as vertical lines
ggplot(SimParametersClean_long1, aes(x = estimate)) +
  geom_density(fill = "gray80", color = "black") +
  geom_vline(data = fit_values1, aes(xintercept = fit), color = "red", linetype = "dashed", size = 1) +
  facet_wrap(~ parameter, scales = "free", ncol = 2) +
  theme_minimal() +
  labs(title = "Density of Simulated Parameters with MLEs", x = "Estimate", y = "Density")

# Histogram plot with fitted values as vertical lines
ggplot(SimParametersClean_long1, aes(x = estimate)) +
  geom_histogram(bins = 5, fill = "gray80", color = "black") +
  geom_vline(data = fit_values1, aes(xintercept = fit), color = "red", linetype = "dashed", size = 1) +
  facet_wrap(~ parameter, scales = "free", ncol = 2) +
  theme_minimal() +
  labs(title = "Histogram of Simulated Parameters with MLEs", x = "Estimate", y = "Count")
```

```{r}
param_labs = c(
  eta   = expression(eta),
  a     = "a",
  c     = "c",
  gamma = expression(gamma)
)

# Density plot
p_density_M1 = ggplot(SimParametersClean_long1, aes(x = estimate)) +
  geom_density(
    fill = "grey85",
    color = "grey25",
    linewidth = 0.6,
    alpha = 0.8
  ) +
  geom_vline(
    data = fit_values1,
    aes(xintercept = fit),
    color = "red3",
    linetype = "dashed",
    linewidth = 0.8
  ) +
  facet_wrap(
    ~ parameter,
    scales = "free",
    ncol = 2,
    labeller = as_labeller(param_labs, default = label_parsed)
  ) +
  labs(
    title = "Simulation distributions of parameter estimates: true parameter estimated by Model 1",
    subtitle = "Dashed red lines show fitted MLEs Model 1 from the observed data",
    x = "Estimate",
    y = "Density"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.caption = element_text(hjust = 0.5, color = "grey35"),
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )


p_hist_M1 = ggplot(SimParametersClean_long1, aes(x = estimate)) +
  geom_histogram(
    bins = 15,
    fill = "grey85",
    color = "grey25",
    linewidth = 0.3
  ) +
  geom_vline(
    data = fit_values1,
    aes(xintercept = fit),
    color = "red3",
    linetype = "dashed",
    linewidth = 0.8
  ) +
  facet_wrap(
    ~ parameter,
    scales = "free",
    ncol = 2,
    labeller = as_labeller(param_labs, default = label_parsed)
  ) +
  labs(
    title = "Simulation histograms of parameter estimates: true parameter estimated by Model 1",
    subtitle = "Dashed red lines show fitted MLEs Model 1 from the observed data",
    x = "Estimate",
    y = "Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.caption = element_text(hjust = 0.5, color = "grey35"),
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

# Testing for Grid Starting Values

```{r}
theta_grid = expand.grid(
  eta   = c(60, 70),
  a     = c(-130, -150),
  b     = c(0, 10),
  c     = c(100, 115),
  d     = c(0, 10),
  gamma = c(8000, 6000)
)

fits_grid = vector("list", nrow(theta_grid))

for (j in 1:nrow(theta_grid)) {
  
  theta_start_j = as.numeric(theta_grid[j, ])
  names(theta_start_j) = names(theta_grid)
  
  cat("Running start", j, "of", nrow(theta_grid), "\n")
  
  fitBFGS_j = optim(
        par = theta_start_j,
        fn = neg_loglik,
        data = Ca2_data,
        h = 0.02,
        method = "L-BFGS-B",
        lower = c(
          eta = 1e-8,
          a = -Inf,
          b = -Inf,
          c = -Inf,
          d = -Inf,
          gamma = 200
        ),
        upper = c(
          eta = Inf,
          a = -1e-8,
          b = Inf,
          c = Inf,
          d = Inf,
          gamma = Inf
        ))
  
   fits_grid[[j]] = data.frame(
    run = j,
    
    eta_start = theta_start_j["eta"],
    a_start = theta_start_j["a"],
    b_start = theta_start_j["b"],
    c_start = theta_start_j["c"],
    d_start = theta_start_j["d"],
    gamma_start = theta_start_j["gamma"],
    
    eta_hat = fitBFGS_j$par["eta"],
    a_hat = fitBFGS_j$par["a"],
    b_hat = fitBFGS_j$par["b"],
    c_hat = fitBFGS_j$par["c"],
    d_hat = fitBFGS_j$par["d"],
    gamma_hat = fitBFGS_j$par["gamma"],
    
    value = fitBFGS_j$value,
    convergence = fitBFGS_j$convergence
  )
}

grid_results = do.call(rbind, fits_grid)

grid_results = grid_results[order(grid_results$value), ]

head(grid_results)
```

```{r}
grid_long = grid_results %>%
  select(run, convergence, value, ends_with("_hat")) %>%
  pivot_longer(
    cols = ends_with("_hat"),
    names_to = "parameter",
    values_to = "estimate"
  ) %>%
  mutate(
    parameter = gsub("_hat", "", parameter),
    convergence = factor(convergence)
  )

ggplot(grid_long, aes(x = run, y = estimate, colour = convergence)) +
  geom_point(size = 2) +
  geom_line(alpha = 0.4) +
  facet_wrap(~ parameter, scales = "free_y") +
  labs(
    title = "Estimated parameters across starting values",
    x = "Starting value run",
    y = "Estimated parameter",
    colour = "Convergence"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

```{r}
ggplot(grid_long, aes(x = value, y = estimate, colour = convergence)) +
  geom_point(size = 2) +
  facet_wrap(~ parameter, scales = "free_y") +
  labs(
    title = "Estimated parameters versus negative log-likelihood",
    x = "Negative log-likelihood",
    y = "Estimated parameter"
  ) +
  theme_minimal(base_size = 13) 
```

# Bootstrap likelihood

Simulating data with M0 and then fitting both

```{r}

set.seed(16)
# -----------------------------
# Settings
# -----------------------------

M = 150          
h_sim = 0.0001
subselect = 200
h_fit = h_sim * subselect

n_obs = nrow(Ca2_final)

# Observed LR statistic
LR_obs = 2 * (fitBFGS$value - fitBFGS2$value)

# Fitted null parameters: M0
theta_M0_hat = fitBFGS$par

# Starting values
theta_start_M0 = fitBFGS$par
theta_start_M2 = fitBFGS2$par

# Storage
boot_results = vector("list", 38)

for (m in 1:38) {
  
  cat("Bootstrap run", m, "of", M, "\n")
  
  # -----------------------------
  # 1. Simulate under null model M0
  # -----------------------------
  
  SimDataBoot = SimulateModel0_sub(theta = theta_M0_hat, h_sim = h_sim, n_data = 2499, subselect = 200)
  
  print('Simulation finished')

  # -----------------------------
  # 2. Fit restricted model M0
  # -----------------------------
  
  fit_M0_boot =  optim(
      par = theta_start_M0,
      fn = neg_loglik,
      data = SimDataBoot,
      h = h_fit,
      method = "L-BFGS-B",
      lower = c(
        eta = 1e-8,
        a = -Inf,
        b = -Inf,
        c = -Inf,
        d = -Inf,
        gamma = 200
      ),
      upper = c(
        eta = Inf,
        a = -1e-8,
        b = Inf,
        c = Inf,
        d = Inf,
        gamma = Inf
      )
    )
  
  print("M0 finished")
  
  # -----------------------------
  # 3. Fit full model M2
  # -----------------------------
  
  fit_M2_boot =  optim(
      par = theta_start_M2,
      fn = neg_loglik2,
      data = SimDataBoot,
      h = h_fit,
      method = "L-BFGS-B",
      lower = c(
        eta = 1e-8,
        a = -Inf,
        b = -Inf,
        c = -Inf,
        d = -Inf,
        alpha = 1e-8,
        beta = -Inf,
        gamma = 200
      ),
      upper = c(
        eta = Inf,
        a = -1e-8,
        b = Inf,
        c = Inf,
        d = Inf,
        alpha = Inf,
        beta = Inf,
        gamma = Inf
      )
    )
  
  print("M2 finished")
  
  # -----------------------------
  # 4. Save bootstrap LR
  # -----------------------------
  
    nll_M0_boot = fit_M0_boot$value
    nll_M2_boot = fit_M2_boot$value
    
    LR_boot = 2 * (nll_M0_boot - nll_M2_boot)
    
    boot_results[[m]] = data.frame(
      m = m,
      nll_M0 = nll_M0_boot,
      nll_M2 = nll_M2_boot,
      LR = LR_boot,
      conv_M0 = fit_M0_boot$convergence,
      conv_M2 = fit_M2_boot$convergence,
      
      eta_M0 = fit_M0_boot$par["eta"],
      a_M0 = fit_M0_boot$par["a"],
      b_M0 = fit_M0_boot$par["b"],
      c_M0 = fit_M0_boot$par["c"],
      d_M0 = fit_M0_boot$par["d"],
      gamma_M0 = fit_M0_boot$par["gamma"],
      
      eta_M2 = fit_M2_boot$par["eta"],
      a_M2 = fit_M2_boot$par["a"],
      b_M2 = fit_M2_boot$par["b"],
      c_M2 = fit_M2_boot$par["c"],
      d_M2 = fit_M2_boot$par["d"],
      alpha_M2 = fit_M2_boot$par["alpha"],
      beta_M2 = fit_M2_boot$par["beta"],
      gamma_M2 = fit_M2_boot$par["gamma"]
    )
    
}
```

```{r}

combined_boot_results = bind_rows(boot_results)

head(combined_boot_results)
```

```{r}
set.seed(16)
# -----------------------------
# Settings
# -----------------------------

M = 500          # increase later, e.g. 500 or 1000
h_sim = 0.0001
subselect = 200
h_fit = h_sim * subselect

n_obs = nrow(Ca2_final)

# Observed LR statistic
LR_obs = 2 * (fitBFGS$value - fitBFGS2$value)

# Fitted null parameters: M0
theta_M1_hat = fitBFGS1$par

# Starting values
theta_start_M0 = fitBFGS$par
theta_start_M1 = fitBFGS1$par

# Storage
boot_results = vector("list", 38)

for (m in 1:38) {
  
  cat("Bootstrap run", m, "of", M, "\n")
  
  # -----------------------------
  # 1. Simulate under null model M0
  # -----------------------------
  
  SimDataBoot = SimulateModel0_sub(theta = theta_M0_hat, h_sim = h_sim, n_data = 2499, subselect = 200)
  
  print('Simulation finished')

  # -----------------------------
  # 2. Fit restricted model M0
  # -----------------------------
  
  fit_M0_boot =  optim(
      par = theta_start_M0,
      fn = neg_loglik,
      data = SimDataBoot,
      h = h_fit,
      method = "L-BFGS-B",
      lower = c(
        eta = 1e-8,
        a = -Inf,
        b = -Inf,
        c = -Inf,
        d = -Inf,
        gamma = 200
      ),
      upper = c(
        eta = Inf,
        a = -1e-8,
        b = Inf,
        c = Inf,
        d = Inf,
        gamma = Inf
      )
    )
  
  print("M0 finished")
  
  # -----------------------------
  # 3. Fit full model M2
  # -----------------------------
  
  fit_M2_boot =  optim(
      par = theta_start_M2,
      fn = neg_loglik2,
      data = SimDataBoot,
      h = h_fit,
      method = "L-BFGS-B",
      lower = c(
        eta = 1e-8,
        a = -Inf,
        b = -Inf,
        c = -Inf,
        d = -Inf,
        alpha = 1e-8,
        beta = -Inf,
        gamma = 200
      ),
      upper = c(
        eta = Inf,
        a = -1e-8,
        b = Inf,
        c = Inf,
        d = Inf,
        alpha = Inf,
        beta = Inf,
        gamma = Inf
      )
    )
  
  print("M2 finished")
  
  # -----------------------------
  # 4. Save bootstrap LR
  # -----------------------------
  
    nll_M0_boot = fit_M0_boot$value
    nll_M2_boot = fit_M2_boot$value
    
    LR_boot = 2 * (nll_M0_boot - nll_M2_boot)
    
    boot_results[[m]] = data.frame(
      m = m,
      nll_M0 = nll_M0_boot,
      nll_M2 = nll_M2_boot,
      LR = LR_boot,
      conv_M0 = fit_M0_boot$convergence,
      conv_M2 = fit_M2_boot$convergence,
      
      eta_M0 = fit_M0_boot$par["eta"],
      a_M0 = fit_M0_boot$par["a"],
      b_M0 = fit_M0_boot$par["b"],
      c_M0 = fit_M0_boot$par["c"],
      d_M0 = fit_M0_boot$par["d"],
      gamma_M0 = fit_M0_boot$par["gamma"],
      
      eta_M2 = fit_M2_boot$par["eta"],
      a_M2 = fit_M2_boot$par["a"],
      b_M2 = fit_M2_boot$par["b"],
      c_M2 = fit_M2_boot$par["c"],
      d_M2 = fit_M2_boot$par["d"],
      alpha_M2 = fit_M2_boot$par["alpha"],
      beta_M2 = fit_M2_boot$par["beta"],
      gamma_M2 = fit_M2_boot$par["gamma"]
    )
    
}
```

```{r}
n_obs = nrow(Ca2_final)

# Observed LR statistic
LR_obs = 2 * (fitBFGS1$value - fitBFGS$value)

# Fitted null parameters: M0
theta_M1_hat = fitBFGS1$par

# Starting values
theta_start_M0 = fitBFGS$par
theta_start_M1 = fitBFGS1$par

# Storage
BootResultsM0M1 = vector("list", 38)

set.seed(40) 
for (m in 1:M) {
  
  cat("Bootstrap run", m, "of", M, "\n")
  
  # -----------------------------
  # 1. Simulate under null model M0
  # -----------------------------
  
  SimDataBoot = SimulateModel1_sub(theta = theta_M1_hat, h_sim = h_sim, n_data = 2499, subselect = 200)
  
  print('Simulation finished')

  # -----------------------------
  # 2. Fit restricted model M0
  # -----------------------------
  
  fit_M0_boot =  optim(
      par = theta_start_M0,
      fn = neg_loglik,
      data = SimDataBoot,
      h = h_fit,
      method = "L-BFGS-B",
      lower = c(
        eta = 1e-8,
        a = -Inf,
        b = -Inf,
        c = -Inf,
        d = -Inf,
        gamma = 200
      ),
      upper = c(
        eta = Inf,
        a = -1e-8,
        b = Inf,
        c = Inf,
        d = Inf,
        gamma = Inf
      )
    )
  
  print("M0 finished")
  
  # -----------------------------
  # 3. Fit full model M2
  # -----------------------------
  
  fit_M1_boot =  optim(
      par = theta_start_M1,
      fn = neg_loglik1,
      data = SimDataBoot,
      h = h_fit,
      method = "L-BFGS-B",
      lower = c(
        eta = 1e-8,
        a = -Inf,
        c = -Inf,
        gamma = 200
      ),
      upper = c(
        eta = Inf,
        a = -1e-8,
        c = Inf,
        gamma = Inf
      )
    )
  
  print("M1 finished")
  
  # -----------------------------
  # 4. Save bootstrap LR
  # -----------------------------
  
    nll_M0_boot = fit_M0_boot$value
    nll_M1_boot = fit_M1_boot$value
    
    LR_boot = 2 * (nll_M1_boot - nll_M0_boot)
    
    BootResultsM0M1[[m]] = data.frame(
      m = m,
      nll_M0 = nll_M0_boot,
      nll_M1 = nll_M1_boot,
      LR = LR_boot,
      conv_M0 = fit_M0_boot$convergence,
      conv_M1 = fit_M1_boot$convergence,
      
      eta_M0 = fit_M0_boot$par["eta"],
      a_M0 = fit_M0_boot$par["a"],
      b_M0 = fit_M0_boot$par["b"],
      c_M0 = fit_M0_boot$par["c"],
      d_M0 = fit_M0_boot$par["d"],
      gamma_M0 = fit_M0_boot$par["gamma"],
      
      eta_M1 = fit_M1_boot$par["eta"],
      a_M1 = fit_M1_boot$par["a"],
      b_M1 = fit_M1_boot$par["b"],
      c_M1 = fit_M1_boot$par["c"],
      d_M1 = fit_M1_boot$par["d"],
      alpha_M1 = fit_M1_boot$par["alpha"],
      beta_M1 = fit_M1_boot$par["beta"],
      gamma_M1 = fit_M1_boot$par["gamma"]
    )
    
}
```

```{r}
boot_results_M0M1  = bind_rows(BootResultsM0M1)

# ---- p-value ----
mean(boot_results_M0M1$LR>3.800)
```
